---
title: "Power Analysis"
format: 
  revealjs:
    height: 800
editor: source
filters:
  - shinylive
slide-number: true
---


## Research Question

```{=html}
<!---- 
######################################################################
## Interactive Statistics Project - Module 2: Statistical Power
##
## We thank the Association for Psychological Science for a small
## teaching grant to help in the development of this software.
##
## Copyright 2025-2026 Jeremy Rappel, Carl F. Falk, Mira Saad &
##  Jens Kreitewolf
##
## This program is free software: you can redistribute it and/or
## modify it under the terms of the GNU Affero General Public License as
## published by the Free Software Foundation, either version 3 of
## the License, or (at your option) any later version.
##
## This program is distributed in the hope that it will be useful,
## but WITHOUT ANY WARRANTY; without even the implied warranty of
## MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
## GNU Affero General Public License for more details.
## <http://www.gnu.org/licenses/>
##
---->
```

```{r}
library(ggplot2)
library(shiny)
library(plotly)
library(tibble)
library(munsell)

theme_minimalism <- function(){
  theme_minimal() + # ggplot's minimal theme hides many unnecessary features of plot
  theme(            # make modifications to the theme
    panel.grid.major.y=element_blank(),   # hide major grid for y axis
    panel.grid.minor.y=element_blank(),   # hide minor grid for y axis
    panel.grid.major.x=element_blank(),   # hide major grid for x axis
    panel.grid.minor.x=element_blank(),   # hide minor grid for x axis
    text=element_text(size=14),           # font aesthetics 
    axis.text=element_text(size=12),
    axis.title=element_text(size=14,face="bold"))
}
```


::: {.fragment .fade-in fragment-index="1"}
Let's say we're interested in studying the effect of meditation on well-being.
:::

::: {.fragment .fade-in fragment-index="2"}
We want to test if people who meditate have different well-being scores than the non-meditating population.
:::

## The Null Distribution {.smaller}

::: r-stack
```{r}
#| fig-align: center
plot1 <- ggplot(data=data.frame(x=c(35:65)),aes(x)) +
  stat_function(fun=dnorm,
                args=list(mean=50, sd=5),
                geom="area", 
                linetype="solid",
                size=1.25,
                fill=NA,
                aes(color="Null Distribution"),
                xlim=c(35,65)) +
  scale_y_continuous(breaks=NULL) +
  ylab("")+
  xlab("Well-Being Score")+
  scale_colour_manual("Legend",values = c("#b2df8a")) +
  theme_minimalism() 
ggplotly(plot1) |> layout(legend = list(x = 0.9, y = 0.9))
```
:::

::: {.fragment .fade-in fragment-index="1"}
Let's say this is the sampling distribution of the non-meditating population (the **null sampling distribution**) for n=15.
:::

::: {.fragment .fade-in fragment-index="2"}
Let's say this sampling distribution has a mean score of 50 and a standard error of 5.
:::

## The Null Distribution {.smaller}

::: r-stack
```{r}
#| fig-align: center
ggplotly(plot1 +
    stat_function(fun=dnorm,
                  args=list(mean=50.3, sd=5),
                  geom="area", 
                  linetype="dotted", 
                  size=1.25,
                  fill=NA, 
                  aes(color="Alternative Distribution"), 
                  xlim=c(35,65)) +
    scale_colour_manual("Legend",values = c("#a6cee3","#b2df8a"))) |> layout(legend = list(x = 0.9, y = 0.9))
```
:::

::: {.fragment .fade-in fragment-index="1"}
If the null hypothesis (H~0~) were true (If meditation had no effect on well-being), the sampling distribution of the meditation (i.e. experimental) group would be identical to the non-meditation (i.e. control) group
:::

## The Alternative Distribution {.smaller}

::: r-stack
```{r}
#| fig-align: center
ggplotly(plot2 <- ggplot(data=data.frame(x=c(35:75)),aes(x)) +
  stat_function(fun=dnorm,
                args=list(mean=50, sd=5),
                geom="area", 
                linetype="solid", 
                fill=NA, 
                size=1.25,
                aes(color="Null Distribution"), 
                xlim=c(35,75)) +
  stat_function(fun=dnorm,
                args=list(mean=55, sd=5),
                geom="area", 
                linetype="dotted", 
                size=1.25,
                fill=NA, 
                aes(color="Alternative Distribution"), 
                xlim=c(35,75)) +
  scale_x_continuous(limits=c(35,75),
                     breaks= seq(40,70, by=10)) +
  scale_y_continuous(breaks=NULL) +
  ylab("")+
  xlab("Well-Being Score")+
  scale_colour_manual("Legend",values = c("#a6cee3","#b2df8a")) +
  theme_minimalism()) |> layout(legend = list(x = 0.9, y = 0.9))
```
:::

::: {.fragment .fade-in fragment-index="1"}
But, if the people who meditate have higher (or lower) well-being scores than the non-meditating population, then the alternative hypothesis (H~1~) would be true, and the two sampling distributions would be different.
:::

:::: r-stack
::: {.fragment .fade-in fragment-index="2"}
H~0~: μ~Meditating~ = 0
:::
::::

:::: r-stack
::: {.fragment .fade-in fragment-index="2"}
H~1~: μ~Meditating~ ≠ 0
:::
::::

## Sampling {.smaller}

::: r-stack
```{r}
#| fig-align: center
plot3 <- plot2 +
    stat_function(fun=dnorm, # The lower critical region
                args=list(mean=50, sd=5),
                geom="area",
                linetype="solid",
                fill="#1f78b4", 
                aes(color="Critical Region"),
                xlim= c(35,qnorm(.025,mean=50,sd=5))) +
  stat_function(fun=dnorm, # The upper critical region
                args=list(mean=50, sd=5),
                geom="area",
                linetype="solid",
                fill="#1f78b4", 
                aes(color="Critical Region"),
                xlim= c(qnorm(.025, mean=50,sd=5, lower.tail=F),75)) +
    scale_colour_manual("Legend",values = c("Null Distribution" = "#b2df8a","Alternative Distribution"= "#a6cee3","Critical Region" ="#1f78b4"))
ggplotly(plot3) |> layout(legend = list(x = 0.9, y = 0.9))
```
:::

::: {.fragment .fade-in fragment-index="1"}
Let's say the population mean of the meditation (i.e. experimental) group is 5 points higher than the population mean of the non-meditating (i.e. control) group.
:::

::: {.fragment .fade-in fragment-index="2"}
To reject H~0~, our sample mean would need to fall within the critical region of the null distribution (here, alpha=.05, two-tailed).
:::

## Sampling {.smaller}

::: r-stack
```{r}
#| fig-align: center
ggplotly(plot3 +
  geom_vline(xintercept=57, alpha=.5) + 
  annotate("text",x=57,y=.05, label=paste0("Sample mean = 57"))) |> layout(legend = list(x = 0.9, y = 0.9))
```
:::

::: {.fragment .fade-in fragment-index="1"}
Let's say we sample from the meditating population and get a sample mean of 57.
:::

## Sampling {.smaller}

::: {.fragment .fade-in fragment-index="1"}
In this case, our sample mean does not fall within the critical region of the null distribution, so we would fail to reject H0.
:::

::: {.fragment .fade-in fragment-index="2"}
But, there is a true difference between the groups in the population.
:::

::: {.fragment .fade-in fragment-index="3"}
This is an example of a false negative, or type II error.
:::

## Statistical Power

::: r-stack
```{r}
#| fig-align: center
ggplotly(plot3 +
    stat_function(fun=dnorm, # beta overlay
                args=list(mean=55, sd=5),
                geom="area",
                fill="#fb9a99", 
                aes(color="Beta"),
                alpha=.5,
                xlim= c(qnorm(.025, mean=50,sd=5),qnorm(.025, mean=50,sd=5, lower.tail=F)))+
      scale_colour_manual("Legend",values = c("Null Distribution" = "#b2df8a",
                                              "Alternative Distribution"= "#a6cee3",
                                              "Critical Region" ="#1f78b4",
                                              "Beta" = "#fb9a99"))) |> layout(legend = list(x = 0.9, y = 0.9))
```
:::

::: {.fragment .fade-in fragment-index="1"}
The region of the alternative distribution that overlaps with the non-critical region of the null distribution is *beta* (β).
:::

::: {.fragment .fade-in fragment-index="2"}
Sometimes this is called the *false negative rate*, or the probability of getting a Type II error.
:::

## Statistical Power {.smaller}

::: r-stack
```{r}
#| fig-align: center
ggplotly(plot3 +
      stat_function(fun=dnorm, # power overlay
                args=list(mean=55, sd=5),
                geom="area",
                fill="#33a02c", 
                aes(color="Power"),
                alpha=.5,
                xlim= c(qnorm(.025, mean=50,sd=5,lower.tail=F),75))+
      scale_colour_manual("Legend",values = c("Null Distribution" = "#b2df8a",
                                              "Alternative Distribution"= "#a6cee3",
                                              "Critical Region" ="#1f78b4",
                                              "Power" = "#33a02c"))) |> layout(legend = list(x = 0.9, y = 0.9))
```
:::

::: {.fragment .fade-in fragment-index="1"}
1 -β , then, is our *statistical power*. Sometimes it's called the *true positive rate*.
:::

::: {.fragment .fade-in fragment-index="2"}
We can think of it along of the lines of "if we sampled from the population, what's the probability of us (correctly) rejecting H~0~?"
:::

## A Special Case: H~0~ is true {.smaller}

::: r-stack
```{r}
#| fig-align: center
ggplotly(plot4 <- ggplot(data=data.frame(x=c(35:65)),aes(x)) +
  stat_function(fun=dnorm,
                args=list(mean=50, sd=5),
                geom="area", 
                linetype="solid", 
                fill=NA, 
                size=1.25,
                aes(color="Null Distribution"), 
                xlim=c(35,65)) +
  stat_function(fun=dnorm,
                args=list(mean=50, sd=5),
                geom="area", 
                linetype="dotted", 
                size=1.25,
                fill=NA, 
                aes(color="Alternative Distribution"), 
                xlim=c(35,65)) +
        stat_function(fun=dnorm, # The lower critical region
                args=list(mean=50, sd=5),
                geom="area",
                linetype="solid",
                fill="#1f78b4", 
                aes(color="Critical Region"),
                xlim= c(35,qnorm(.025,mean=50,sd=5))) +
  stat_function(fun=dnorm, # The upper critical region
                args=list(mean=50, sd=5),
                geom="area",
                linetype="solid",
                fill="#1f78b4", 
                aes(color="Critical Region"),
                xlim= c(qnorm(.025, mean=50,sd=5, lower.tail=F),65)) +
          stat_function(fun=dnorm, # power overlay (lower)
                args=list(mean=50, sd=5),
                geom="area",
                fill="#33a02c", 
                aes(color="Power"),
                alpha=.5,
                xlim= c(qnorm(.025, mean=50,sd=5,lower.tail=F),65))+
          stat_function(fun=dnorm, #The Power Distribution (Lower)
                args=list(mean=50, sd=5),
                geom="area", 
                #linetype="dotted",
                aes(color="Power"),
                fill="#33a02c",
                alpha=.5,
                xlim=c(35,qnorm(.25))) +
  scale_x_continuous(limits=c(35,65),
                     breaks= seq(40,60, by=10)) +
  scale_y_continuous(breaks=NULL) +
  ylab("")+
  xlab("Well-Being Score")+
    theme_minimalism() + 
        scale_colour_manual("Legend",values = c("Null Distribution" = "#b2df8a",
                                              "Alternative Distribution"= "#a6cee3",
                                              "Critical Region" ="#1f78b4",
                                              "Power" = "#33a02c"))) |> layout(legend = list(x = 0.9, y = 0.9)) 
```
:::

::: {.fragment .fade-in fragment-index="1"}
However, if H~0~ is true, then things change. We can't *correctly* reject H~0~, but we can still get a significant result if our sample mean falls within the critical region.
:::

::: {.fragment .fade-in fragment-index="2"}
This would be a false positive, or an incorrect rejection.
:::

::: {.fragment .fade-in fragment-index="3"}
Because the size of the critical region is determined by α, we sometimes call α the *False Positive Rate*.
:::

## Possible Outcomes {.smaller}

::: {.fragment .fade-in fragment-index="1"}
Thus, we have 4 possible outcomes of a null-hypothesis significance test:
:::

::: {.fragment .fade-in fragment-index="2"}
|   | H~0~ is True \n (in the Population) | H~0~ is False \n (In the population) |
|----|----|---:|
| Retain H~0~ | Correct Retention | False Negative (Type II Error) |
| Reject H~0~ | False Positive (Type I error) | Correct Rejection |
:::

::: {.fragment .fade-in fragment-index="3"}
We can represent the probabilities of each of these outcomes with α and β

|   | H~0~ is True \n (in the Population) | H~0~ is False \n (In the population) |
|----|----|---:|
| Retain H~0~ | 1-α | β |
| Reject H~0~ | α | 1-β |
:::

## Two-Tailed Vs. One-Tailed {.smaller}

The prior examples were for a two-tailed test. What happens if we change to a one-tailed test?

```{shinylive-r}
#| standalone: true
#| panel: fill
#| fig-align: center
#| viewerHeight: 500

library(tibble)
library(munsell)
library(shiny)
library(ggplot2)
library(plotly)

ui <- fluidPage(
  plotlyOutput("tailsPlot"),
  radioButtons("tails", "Tails:",
               c("2 Tails" = "twotail",
                 "1 Tail (Upper)" = "onetail_upper",
                 "1 Tail (Lower)" = "onetail_lower"), inline=T)
)

server <- function(input, output) {
  theme_minimalism <- function(){
    theme_minimal() + # ggplot's minimal theme hides many unnecessary features of plot
    theme(            # make modifications to the theme
      panel.grid.major.y=element_blank(),   # hide major grid for y axis
      panel.grid.minor.y=element_blank(),   # hide minor grid for y axis
      panel.grid.major.x=element_blank(),   # hide major grid for x axis
      panel.grid.minor.x=element_blank(),   # hide minor grid for x axis
      text=element_text(size=14),           # font aesthetics 
      axis.text=element_text(size=12),
      axis.title=element_text(size=14,face="bold"))
  }
  
  output$tailsPlot <- renderPlotly({
      lower <- -4
      upper <- 6
      alpha <- .05
      mu <- 3
      beta <- reactive({    switch(
      input$tails,
        "onetail_upper" = {pnorm(-qnorm(alpha,mean=mu))},
        "onetail_lower" = {pnorm(qnorm(alpha,mean=mu,lower.tail=F))},
        "twotail" = {pnorm(-qnorm(alpha/2,mean=mu)) - pnorm(qnorm(alpha/2,mean=-mu))})})
      power <- (1-beta())
  
      plt <- ggplot(data=data.frame(x=c(lower,upper)),aes(x)) +
        #The true sampling distribution
        stat_function(fun=dnorm,
                      args=list(mean=mu, sd=1),
                      geom="area", 
                      linetype="dotted",
                      size=1.25,
                      fill=NA, 
                      color="#a6cee3",
                      xlim=c(lower,upper))+ 
        #The null distribution
        stat_function(fun=dnorm,
                      args=list(mean=0, sd=1),
                      geom="area",
                      linetype="solid",
                      fill=NA, 
                      size=1.25,
                      color="#b2df8a",
                      xlim=c(lower,upper))+ 
        #Displays the (upper) critical region of the null distribution
        stat_function(fun=dnorm,
                      args=list(mean=0, sd=1),
                      geom="area",
                      fill="#1f78b4", 
                      color="#1f78b4",
                      xlim=switch(
                        input$tails,
                        "onetail_upper" = {c(qnorm(alpha,lower.tail=F), upper)},
                        "onetail_lower" = {c(lower, qnorm(alpha))},
                        "twotail" = {c(qnorm(alpha/2, lower.tail=F),upper)}
                        )) +   
        #Displays the (lower) critical region of the null distribution if two-tailed is selected
        # stat_function(fun=dnorm,
        #               args=list(mean=0, sd=1),
        #               geom="area",
        #               fill="#1f78b4", 
        #               color="#1f78b4",
        #               xlim=switch(
        #                 input$tails,
        #                 "twotail" = {c(lower,qnorm(alpha/2))})) +
                      #xlim= if (input$tails == "twotail") {c(lower,qnorm(alpha/2))} else {c(0,0)}) +
        #Displays the power overlay
              stat_function(fun=dnorm, 
                      args=list(mean=mu, sd=1),
                      geom="area",
                      fill="#33a02c", 
                      color="#33a02c", 
                      alpha=.5,
                      xlim=if (input$tails == "twotail") {
                        c(qnorm(alpha/2,lower.tail=F),upper)} 
                      else if (input$tails == "onetail_upper") {
                        c(qnorm(alpha,lower.tail=F),upper)} 
                      else {
                        c(lower,qnorm(alpha))}) +
        #annotate("text",x=Inf,y=Inf, label = paste0("Power = ", round(power,3)), vjust = 1, hjust = 1) +
        annotate("text",x=4,y=.5, label = paste0("Power = ", round(power,3)), vjust = 1, hjust = 1) +
        scale_x_continuous(limits=c(lower,upper),breaks= seq(round(lower),round(upper),by=1)) +
        #Clears the y-axis ticks
        scale_y_continuous(breaks=NULL) +
          ylab("")+
      xlab("Well-Being Score")+
    theme_minimalism()
      
      if (input$tails == "twotail") {
        plt + stat_function(fun=dnorm,
                      args=list(mean=0, sd=1),
                      geom="area",
                      fill="#1f78b4", 
                      color="#1f78b4",
                      xlim={c(c(lower,qnorm(alpha/2)))})}
      else {plt}
      
  })
}

shinyApp(ui, server)
```

## Power

Statistical power is affected by:

::: {.fragment .fade-in fragment-index="1"}
-   Our (population) effect size
:::

::: {.fragment .fade-in fragment-index="2"}
-   Our sample size
:::

::: {.fragment .fade-in fragment-index="3"}
-   Our chosen alpha value
:::

## (Population) Effect Size {.smaller}

```{shinylive-r}
#| standalone: true
#| panel: fill
#| fig-align: center
#| viewerHeight: 500

library(tibble)
library(munsell)
library(shiny)
library(ggplot2)
library(plotly)

ui <- fluidPage(
  plotlyOutput(outputId = "effectsizePlot"),
  sliderInput("mu", "Effect Size:",min=-4,max=4,value=1,step=.5)
)

server <- function(input, output) {
  theme_minimalism <- function(){
    theme_minimal() + # ggplot's minimal theme hides many unnecessary features of plot
    theme(            # make modifications to the theme
      panel.grid.major.y=element_blank(),   # hide major grid for y axis
      panel.grid.minor.y=element_blank(),   # hide minor grid for y axis
      panel.grid.major.x=element_blank(),   # hide major grid for x axis
      panel.grid.minor.x=element_blank(),   # hide minor grid for x axis
      text=element_text(size=14),           # font aesthetics 
      axis.text=element_text(size=12),
      axis.title=element_text(size=14,face="bold"))
  }
  
  output$effectsizePlot <- renderPlotly({
      lower <- -8
      upper <- 8
      mu <- 5
      beta_value <- reactive({pnorm(-qnorm(.025,mean=input$mu)) - pnorm(qnorm(.025,mean=-input$mu))})
      power <- 1-beta_value()
      stdev <- 5
      se <- stdev/sqrt(25)
      alpha <- .05
  effectsize <- ggplot(data=data.frame(x=c(lower,upper)),aes(x)) +
    stat_function(fun=dnorm, #The Null Distribution
                  args=list(mean=0, sd=1),
                  geom="area", 
                  linetype="solid",
                  fill=NA, 
                  size=1.25,
                  color="#b2df8a",
                  xlim=c(lower,upper)) +
    stat_function(fun=dnorm, # The lower critical region
                  args=list(mean=0, sd=se),
                  geom="area",
                  fill="#1f78b4", 
                  color="#1f78b4",
                  xlim= {c(lower,qnorm(.025))}) +
    stat_function(fun=dnorm, # The upper critical region
                  args=list(mean=0, sd=se),
                  geom="area",
                  fill="#1f78b4", 
                  color="#1f78b4",
                  xlim= {c(qnorm(.025, lower.tail=F),upper)}) +
    stat_function(fun=dnorm, #The Alternative Distribution
                  args=list(mean=input$mu, sd=1),
                  geom="area", 
                  linetype="dotted",
                  size=1.25,
                  fill=NA, 
                  color="#a6cee3",
                  xlim=c(lower,upper)) +
    stat_function(fun=dnorm, #The Power Distribution (Upper)
                  args=list(mean=input$mu, sd=1),
                  geom="area", 
                  #linetype="dotted",
                  color="#33a02c",
                  fill="#33a02c",
                  alpha=.5,
                  xlim=c(qnorm(alpha/2,lower.tail=F),upper)) +
        stat_function(fun=dnorm, #The Power Distribution (Lower)
                  args=list(mean=input$mu, sd=1),
                  geom="area", 
                  #linetype="dotted",
                  color="#33a02c",
                  fill="#33a02c",
                  alpha=.5,
                  xlim=c(lower,qnorm(alpha/2))) +
    #annotate("text",x=Inf,y=Inf, label = paste0("Power = ", round(power,3)), vjust = 1, hjust = 1) +
    annotate("text",
             x=4,
             y=.5, 
             label = paste0("Power = ", round(power,3)),
             vjust = 1, 
             hjust = 1) +
    #Clears the y-axis label
    ylab("")+
    xlab("Well-Being Score")+
    #Sets the x axis ticks to cover the whole plot
    scale_x_continuous(limits=c(lower,upper),breaks= seq(round(lower),round(upper),by=1)) +
    #Clears the y-axis ticks
    scale_y_continuous(breaks=NULL) +
    theme_minimalism()
    if(input$mu == 0) {
      ggplotly(effectsize + annotate("text",x=0,y=.35, label = paste0("Note: When H0 is true, power no longer represents the \n chance of correctly rejecting H0 (because H0 is true, \n we can’t correctly reject it). Instead, it becomes our false-positive rate – the \n chance of incorrectly rejecting H0 (note this is our alpha value)."), size = 14/.pt))
    } else {
      ggplotly(effectsize)
    }
  
  })
}

shinyApp(ui, server)
```

In this case, the effect size represents how far away the means of the two distributions are, in standard deviations.

<!-- (or, how much of a difference there is between the meditating and non-meditating populations). -->

Try changing the effect size. What happens to power as the absolute value of the effect size increases?

## Sample

```{shinylive-r}
#| standalone: true
#| panel: fill
#| fig-align: center
#| viewerHeight: 550

library(tibble)
library(munsell)
library(shiny)
library(ggplot2)
library(plotly)

ui <- fluidPage(
  plotlyOutput("samplesizePlot"),
  sliderInput("sampSize", "Sample Size:",min=2,max=50,value=25,step=1)
)

server <- function(input, output) {
  theme_minimalism <- function(){
    theme_minimal() + # ggplot's minimal theme hides many unnecessary features of plot
    theme(            # make modifications to the theme
      panel.grid.major.y=element_blank(),   # hide major grid for y axis
      panel.grid.minor.y=element_blank(),   # hide minor grid for y axis
      panel.grid.major.x=element_blank(),   # hide major grid for x axis
      panel.grid.minor.x=element_blank(),   # hide minor grid for x axis
      text=element_text(size=14),           # font aesthetics 
      axis.text=element_text(size=12),
      axis.title=element_text(size=14,face="bold"))
  }
  
  output$samplesizePlot <- renderPlotly({
      lower <- -8
      upper <- 8
      mu <- 1
      stdev <- 5
      se <- stdev/sqrt(input$sampSize)
      beta_value <- reactive({pnorm(-qnorm(.025,mean=mu/sqrt(se))) - pnorm(qnorm(.025,mean=-mu/sqrt(se)))})
      power <- 1-beta_value()
      
      alpha <- .05
  ggplot(data=data.frame(x=c(lower,upper)),aes(x)) +
    stat_function(fun=dnorm, #The Null Distribution
                  args=list(mean=0, sd=se),
                  geom="area", 
                  linetype="solid",
                  fill=NA, 
                  size=1.25,
                  color="#b2df8a",
                  xlim=c(lower,upper)) +
    stat_function(fun=dnorm, # The lower critical region
                  args=list(mean=0, sd=se),
                  geom="area",
                  fill="#1f78b4", 
                  color="#1f78b4",
                  xlim= {c(lower,qnorm(pnorm(-1.96*(se))))}) +
    stat_function(fun=dnorm, # The upper critical region
                  args=list(mean=0, sd=se),
                  geom="area",
                  fill="#1f78b4", 
                  color="#1f78b4",
                  xlim= {c(qnorm(pnorm(-1.96*(se)), lower.tail=F),upper)}) +
    stat_function(fun=dnorm, #The Alternative Distribution
                  args=list(mean=mu, sd=se),
                  geom="area", 
                  linetype="dotted",
                  fill=NA, 
                  size=1.25,
                  color="#a6cee3",
                  xlim=c(lower,upper)) +
    stat_function(fun=dnorm, #The Power Distribution
                  args=list(mean=mu, sd=se),
                  geom="area", 
                  #linetype="dotted",
                  color="#33a02c",
                  fill="#33a02c",
                  alpha=.5,
                  xlim=c(qnorm(pnorm(-1.96*(se)), lower.tail=F),upper)) +
    #annotate("text",x=Inf,y=Inf, label = paste0("Power = ", round(power,3)), vjust = 1, hjust = 1) +
    annotate("text",x=4,y=.5, label = paste0("Power = ", round(power,3)), vjust = 1, hjust = 1) +
    #Clears the y-axis label
    ylab("")+
    xlab("Well-Being Score")+
    #Sets the x axis ticks to cover the whole plot
    scale_x_continuous(limits=c(lower,upper),breaks= seq(round(lower),round(upper),by=1)) +
    #scale_y_continuous(limits=c(0,.1)) +
    #Clears the y-axis ticks
    scale_y_continuous(breaks=NULL) +
    theme_minimalism()
  })
}

shinyApp(ui, server)
```

Try changing the sample size. What happens to power as the sample size increases?

## Alpha Value


```{shinylive-r}
#| standalone: true
#| panel: fill
#| fig-align: center
#| viewerHeight: 550

library(tibble)
library(munsell)
library(shiny)
library(ggplot2)
library(plotly)

ui <- fluidPage(
  plotlyOutput("alphaPlot"),
  sliderInput("alpha", "Alpha (α):",min=.01,max=.05,value=.05,step=.01)
)

server <- function(input, output) {
  theme_minimalism <- function(){
    theme_minimal() + # ggplot's minimal theme hides many unnecessary features of plot
    theme(            # make modifications to the theme
      panel.grid.major.y=element_blank(),   # hide major grid for y axis
      panel.grid.minor.y=element_blank(),   # hide minor grid for y axis
      panel.grid.major.x=element_blank(),   # hide major grid for x axis
      panel.grid.minor.x=element_blank(),   # hide minor grid for x axis
      text=element_text(size=14),           # font aesthetics 
      axis.text=element_text(size=12),
      axis.title=element_text(size=14,face="bold"))
  }
  
  output$alphaPlot <- renderPlotly({
    lower <- -8
      upper <- 8
      mu <- 1
      beta_value <- reactive({pnorm(-qnorm(input$alpha/2,mean=mu)) - pnorm(qnorm(input$alpha/2,mean=-mu))})
      power <- 1-beta_value()
      stdev <- 5
      se <- stdev/sqrt(25)
      #sampSize <- 1
      
  ggplot(data=data.frame(x=c(lower,upper)),aes(x)) +
    stat_function(fun=dnorm, #The Null Distribution
                  args=list(mean=0, sd=se),
                  geom="area", 
                  linetype="solid",
                  fill=NA, 
                  size=1.25,
                  color="#b2df8a",
                  xlim=c(lower,upper)) +
      stat_function(fun=dnorm, # The lower critical region
                  args=list(mean=0, sd=se),
                  geom="area",
                  fill="#1f78b4", 
                  color="#1f78b4",
                  xlim= {c(lower,qnorm(input$alpha/2))}) +
    stat_function(fun=dnorm, # The upper critical region
                  args=list(mean=0, sd=se),
                  geom="area",
                  fill="#1f78b4", 
                  color="#1f78b4",
                  xlim= {c(qnorm(input$alpha/2, lower.tail=F),upper)}) +
    stat_function(fun=dnorm, #The Alternative Distribution
                  args=list(mean=mu, sd=se),
                  geom="area", 
                  linetype="dotted",
                  fill=NA, 
                  size=1.25,
                  color="#a6cee3",
                  xlim=c(lower,upper)) +
    stat_function(fun=dnorm, #The Power Distribution
                  args=list(mean=mu, sd=se),
                  geom="area", 
                  #linetype="dotted",
                  color="#33a02c",
                  fill="#33a02c",
                  alpha=.5,
                  xlim=c(qnorm(input$alpha/2,lower.tail=F),upper)) +
    #annotate("text",x=Inf,y=Inf, label = paste0("Power = ", round(power,3)), vjust = 1, hjust = 1) +
    annotate("text",x=4,y=.5, label = paste0("Power = ", round(power,3)), vjust = 1, hjust = 1) +
    #Clears the y-axis label
    ylab("")+
    xlab("Well-Being Score")+
    #Sets the x axis ticks to cover the whole plot
    scale_x_continuous(limits=c(lower,upper),breaks= seq(round(lower),round(upper),by=1)) +
    #Clears the y-axis ticks
    scale_y_continuous(breaks=NULL) +
    theme_minimalism()
    
  })
}

shinyApp(ui, server)
```

Try adjusting the alpha value. What happens to power as alpha decreases?

## Power Analysis for Sample Size [^1]

::: {.fragment .fade-in fragment-index="1"}
A common use for power analysis is to determine what sample size we'll need to find a hypothesized effect before starting a study.
:::

[^1]: Sometimes called *a-priori* power analysis

::: {.fragment .fade-in fragment-index="2"}
Let's say we're planning a study to assess if the well-being scores of recent graduates is different from the general population.
:::


## Power Analysis for Sample Size {.smaller}

Let's say we're planning a study to assess if the well-being scores of recent graduates is different from the general population.

Let's also assume similar research finds a typical effect size of d=.5. Assuming α=.05 (two-tailed), what sample size would we need to have a power of at least .8?

```{shinylive-r}
#| standalone: true
#| panel: fill
#| fig-align: center
#| viewerHeight: 550

library(tibble)
library(munsell)
library(shiny)
library(ggplot2)
library(plotly)
library(bslib)

ui <- fluidPage(
plotlyOutput("aprioriPlot"),
layout_columns(
sliderInput("alpha", "Alpha (α):",min=.01,max=.05,value=.01,step=.01),
sliderInput("effectSize", "Effect Size (d):",min=.05,max=1.5,value=.75,step=.05),
sliderInput("power", "Power (1 - β):",min=.01,max=.99,value=.95,step=.05)))
#p(textOutput("sampleSize"), style="size:16"))

server <- function(input,output) {
  
theme_minimalism <- function(){
  theme_minimal() + # ggplot's minimal theme hides many unnecessary features of plot
  theme(            # make modifications to the theme
    panel.grid.major.y=element_blank(),   # hide major grid for y axis
    panel.grid.minor.y=element_blank(),   # hide minor grid for y axis
    panel.grid.major.x=element_blank(),   # hide major grid for x axis
    panel.grid.minor.x=element_blank(),   # hide minor grid for x axis
    text=element_text(size=14),           # font aesthetics 
    axis.text=element_text(size=12),
    axis.title=element_text(size=14,face="bold"))
}

output$sampleSize <- renderText({samp <- reactive({((qnorm(1-input$alpha/2)+qnorm(input$power))^2)/input$effectSize^2})
  round(samp(),0)})

output$aprioriPlot <- renderPlotly({
  lower <- -3
    upper <- 8
    es <- input$effectSize
    sampleSize <- reactive({((qnorm(1-input$alpha/2)+qnorm(input$power))^2)/input$effectSize^2})
    se <- 1/sqrt(sampleSize())
    tVal <- es/(1/sqrt(sampleSize()))
    #sampSize <- 1
    
ggplot(data=data.frame(x=c(lower,upper)),aes(x)) +
  stat_function(fun=dnorm, #The Null Distribution
                args=list(mean=0, sd=1),
                geom="area", 
                linetype="solid",
                fill=NA, 
                size=1.25,
                color="#b2df8a",
                xlim=c(lower,upper)) +
  stat_function(fun=dnorm, #The Alternative Distribution
                args=list(mean=tVal, sd=1),
                geom="area", 
                linetype="dotted",
                fill=NA, 
                size=1.25,
                color="#a6cee3",
                xlim=c(lower,upper)) +
  stat_function(fun=dnorm, # The lower critical region
                args=list(mean=0, sd=1),
                geom="area",
                fill="#1f78b4", 
                color="#1f78b4",
                xlim= {c(lower,qnorm(input$alpha/2))}) +
  stat_function(fun=dnorm, # The upper critical region
                args=list(mean=0, sd=1),
                geom="area",
                fill="#1f78b4", 
                color="#1f78b4",
                xlim= {c(qnorm(input$alpha/2, lower.tail=F),upper)}) +
   stat_function(fun=dnorm, #The Power Distribution
                args=list(mean=tVal, sd=1),
                geom="area", 
                #linetype="dotted",
                color="#33a02c",
                fill="#33a02c",
                alpha=.5,
                xlim=c(qnorm(input$alpha/2,lower.tail=F),upper)) +
  #annotate("text",x=Inf,y=Inf, label = paste0("Power = ", round(power,3)), vjust = 1, hjust = 1) +
  annotate("text",x=2.5,y=.5, label = paste0("Required Sample Size = ", round(sampleSize(),0)), vjust = 1, hjust = 1, size = unit(7, "pt")) +
  #Clears the y-axis label
  ylab("")+
  xlab("Well-Being Score")+
  #Sets the x axis ticks to cover the whole plot
  scale_x_continuous(limits=c(lower,upper),breaks= seq(round(lower),round(upper),by=1)) +
  #Clears the y-axis ticks
  scale_y_continuous(breaks=NULL) +
  theme_minimalism()
  
})}

shinyApp(ui, server)
```
